---
title: "qPCR data analysis"
author: "Darya Vanichkina"
date: "05/03/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(openxlsx)
library(tidyverse)

```

## Load the data

As the first step, we load the data from an xlsx Excel file using the openxlsx library. 

I am reading each of the sheets I am interested in into variables with meaningful names. I use the startRow parameter to specify that I want to start reading in the data starting from row 36 (i.e. I don't want to read any of the metadata above). 

Note: make sure you use the latest version (openxlsx_4.0.17+) of this library in order to have the same parameters available to you as I have used. 

```{r LoadData}
sample_setup <- readWorkbook("qPCR_sample.xlsx",sheet = "Sample Setup", startRow = 36, colNames = TRUE, rowNames = FALSE, detectDates = FALSE, skipEmptyRows = TRUE,skipEmptyCols = TRUE)

amplification_data <- readWorkbook("qPCR_sample.xlsx",sheet = "Amplification Data", startRow = 36, colNames = TRUE, rowNames = FALSE, detectDates = FALSE, skipEmptyRows = TRUE, na.strings = "",skipEmptyCols = TRUE)

qPCR_results <- readWorkbook("qPCR_sample.xlsx",sheet = "Results", startRow = 36, colNames = TRUE, rowNames = FALSE, detectDates = FALSE, skipEmptyRows = TRUE, skipEmptyCols = TRUE, na.strings = "Undetermined")

sample_annotation <- readWorkbook("qPCR_sample.xlsx",sheet = "Sample Annotation", startRow = 1, colNames = TRUE, rowNames = FALSE, detectDates = FALSE, skipEmptyRows = TRUE, skipEmptyCols = TRUE)

```

Next, I try to reproduce what you would have seen on the qPCR machine - the amplification curves observed for each well. This allows me to see that there is actually one well which is an outlier - it has much higher Rn values even at baseline than all of the other wells. It may be prudent to exclude this well, although for this analysis I won't, since it does demonstrate a similar curve as all the other wells. However, I will show how to identify this well based on what we see in the plot.


```{r PlotAmplif}
amplification_data_complete <- amplification_data[complete.cases(amplification_data),]

# look at all the wells
ggplot(amplification_data_complete, aes(x = Cycle, y = Rn, colour = Well)) + geom_point() + theme_minimal()

# zoom into wells 30:40
amplification_data_complete %>% filter(Well %in% 30:40) %>% ggplot(., aes(x = Cycle, y = Rn, colour = Well)) + geom_point() + theme_minimal()

# zoom into well 30:34, and show it as a curve, not a series of dots
amplification_data_complete %>% filter(Well %in% 30:34) %>% ggplot(., aes(x = Cycle, y = Rn, colour = as.factor(Well))) + geom_path() + theme_minimal() + scale_color_brewer(palette = "Set1")

# identify the outlier - it's well 257, Gene 8
amplification_data_complete %>% filter(Cycle == 1) %>% filter(Rn > 2.5)
amplification_data_complete %>% filter(Well %in% 257) %>% ggplot(., aes(x = Cycle, y = Rn, colour = as.factor(Well))) + geom_path() + theme_minimal() + scale_color_brewer(palette = "Set1")
```
I could have then used subset data frame to exluce this well. 


Next, define the standard functions for quantitating qPCR data using the [delta delta Ct method](http://bitesizebio.com/24894/4-easy-steps-to-analyze-your-qpcr-data-using-double-delta-ct-analysis/), and the error for this method.


```{r DefineFunctions}
deltadeltact <- function(geneofinterestMean , controlMean){
  return(2 ^ - (geneofinterestMean - controlMean))
}

deltadeltact_error <- function(geneofinterestMean, controlMean, geneofinterestSD, controlSD){
  cterror <- 2 ^-(geneofinterestMean - controlMean + sqrt(geneofinterestSD^2 + controlSD^2))
  return(deltadeltact(geneofinterestMean, controlMean) - cterror)
}

```

Now merge the tables that we generated from the different sheets of the spreadsheet:

```{r MergeTables}
qPCR_results_anno <- merge.data.frame(qPCR_results, sample_annotation, by.x = "Sample.Name", by.y = "Sample")
# dplyr syntax
qPCR_results_anno2 <- inner_join(qPCR_results, sample_annotation, by = c("Sample.Name" = "Sample"))  %>% select(`Target.Name`, `siRNA`, `CT`, `Time`)
qPCR_results_anno2 <- qPCR_results_anno2[complete.cases(qPCR_results_anno2),]
```

Filter, do the math, and plot the results.


```{r NowPlot}
qPCR_results_anno2_genes <- qPCR_results_anno2 %>% 
  filter(Target.Name != "ReferenceGene") %>% # select the non-refernce gene values
  unite(id, Target.Name, siRNA, Time) %>%  # paste Target.Name,siRNA and Time columns into one column called id
  group_by(id) %>% # group by this id column
  summarise( mean_genes = mean(CT), sd_genes = sd(CT)) %>% # make new columns that have the mean and standard deviation of CT for the three replicates for each id 
  separate(id,  into = c("Gene", "siRNA", "Time") ,sep = "_") %>% # now separate the id column into Gene, siRNA, Time 
  unite(id, siRNA, Time) # then unite siRNA and Time into a new column called id


# same as above except only looking at control genes
qPCR_results_anno2_control <- qPCR_results_anno2 %>% 
  filter(Target.Name == "ReferenceGene") %>% 
  unite(id, Target.Name, siRNA, Time) %>% 
  group_by(id) %>% 
  summarise( mean_genes = mean(CT), sd_genes = sd(CT)) %>% 
  separate(id,  into = c("Gene", "siRNA", "Time") ,sep = "_") %>% 
  unite(id, siRNA, Time) %>% select(id, mean_control = mean_genes , sd_control = sd_genes)


qPCR_processed_table <- merge.data.frame(qPCR_results_anno2_genes, qPCR_results_anno2_control, by = "id", all.x = TRUE, all.y = TRUE) %>% # merge results for control gene and genes of interest
  separate(id,  into = c("siRNA", "Time") ,sep = "_", remove = FALSE) %>% # separate the id column into siRNA and Time 
  group_by(siRNA, Time, Gene) %>% # group the data by siRNA, Time and Gene
  summarise(ddCt = deltadeltact(mean_genes, mean_control), ddCtError = deltadeltact_error(mean_genes, mean_control, sd_genes, sd_control)) # use the summarise function to apply our custom defined functions deltadeltact and deltadeltact_error to the data grouped by siRNA, time and gene
  

# now plot

qPCR_processed_table %>% 
  ggplot(.,aes(x = siRNA, y = ddCt, fill = siRNA)) + geom_bar(stat = "identity") + facet_grid(Gene ~ Time, scales = "free") + theme_bw() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + geom_errorbar(aes(ymin=ddCt-ddCtError, ymax=ddCt + ddCtError), width=.1,  position=position_dodge(.9))

```


